Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-28 2024pipit/section/30-Core.tex
copy and tighten

For values, we use an auxiliary definition $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
Likewise, for primitives we use $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

@@

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types; the result is the same type.

@@

Rule \textsc{TCheck} states that statically checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee clauses must be boolean expressions.
Additionally, the guarantee clause can refer to the result value as $x$.

@@

At a high level, evaluation proceeds by unfolding the recursive streams until a concrete value is determined.
For example, to evaluate the earlier sum example with input stream $i = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumionetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream to get $(\xfby{0}{\sumrec}) + i$.
To evaluate variables, we look for the variable in the current, or most recent, heap:
$$
\ruleIN{
}{
  \bigstep{\sumionetwo}{i}{2}
}{Var}
$$

@@

Returning to \autoref{f:core-bigstep}, rule {\sc Value} states that evaluating a value results in the value itself.
Rule {\sc Var} states that to evalute a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap, we look up the variable in $\sigma$.
Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
Rule {\sc Let} is standard.

For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value or not.
When there is no previous value -- the streaming history consists of just the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
Otherwise, rule $\textsc{Fby}_S$ applies.
In this case, the rule evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.

@@

Rule {\sc Check} states that check expressions always evaluate to unit.
We do not perform a dynamic check that the property is true here; checking the truth of properties is dealt with in the checked semantics (\autoref{s:core:checked}).

Rule {\sc Contract} states that contracts evaluate by just evaluating their body.
Like with checks, we do not perform a dynamic check that the precondition and postcondition hold.
From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
However, such an abstraction would make evaluation non-deterministic and, for contracts with unsatisfiable postconditions, non-total.
The deterministic and total nature of evaluation is key to our proofs and metatheory.

@@

In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
We call these the \emph{checked} semantics.
Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.

The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties of $e$ with status $\PStatus$ hold.
The property status dictates which properties should be checked and which should be ignored.

To show that an expression $e$'s unknown properties hold, we prove that for all streaming histories $\Sigma$, assuming the valid properties hold ($\semcheck{\Sigma}{\PSValid}{e}$), then the unknown properties ($\semcheck{\Sigma}{\PSUnknown}{e}$) hold.
The assumption here means that we do not have to re-check properties after proving them once.

Contracts involve two proofs: one for the definition and one for the instantiation.
To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), and assuming that the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee always holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.

@@

For example, recall our earlier sum contract, which stated that the sum of strictly positive integers is itself positive:
$$
@let@~\text{sum}~i = \sumcontract
$$
To check the contract definition on a concrete input $i = [1; 2]$, we first evaluate the body:
$$
\bigsteps{\sumionetwo}{\sumrec}{[1; 3]}
$$
We then check that, assuming all inputs are positive, then all results are positive:
$$
\bigstepalways{\sumionetwo}{i > 0} \implies
\bigstepalways{\sgl{i \mapsto 1, \sumvar \mapsto 1}; \sgl{i \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
$$

To prove that the contract definition is valid for all inputs, we perform a similar proof for arbitrary input streams ($\forall \Sigma$).
We assume that the rely is true \emph{at all points} in the stream, and must show that the guarantee is true \emph{at all points}.
Consider if we had instead used the input stream $i = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

Contract instantiations often involve feedback loops, where the input to a contract depends, directly or indirectly, on previous output values of the contract itself.
For example, we can compute the Fibonacci sequence by instantiating the sum contract with its output fed back to it\footnote{The program shown here does not strictly fit within the core grammar, which does not support named functions; in the implementation we use the \fstar{} meta-language's top-level definitions for such definitions.}:
$$
  % @rec@~\fibvar.\, \sumcontractX{\PSUnknown}{(\xfby{1}{\fibvar})}
  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
$$

To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
We describe the mechanisms of this proof in \autoref{s:transition}.

% To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
% Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
% This sort of feedback is necessary for proving properties of mutually-dependent calls.
This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
For example, the non-causal program $@rec@~\textit{nope}.\, \text{sum}~(-\textit{nope})$ depends instantaneously on itself.
The rely requires that the negated result is positive; the guarantee ensures that the result is positive.
However, there is no positive number whose negation is also positive, therefore the rely cannot imply the guarantee.
This potential issue is resolved by outlawing such non-causal programs.

@@

The checked semantics of Pipit is defined in \autoref{f:core-check}.
The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
For binding forms such as let-expressions and recursive-expressions, the semantics evaluates the corresponding expression and extends the streaming history with the bound variable.

@@

Rule {\sc ChkCheck} applies when checking property status $\PStatus$ of an expression $\xcheckP{\PStatus'}{e}$.
If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we perform the actual check by evaluating the expression $e$ and requiring it to evaluate to a stream of trues.
Otherwise, we do not need to evaluate the check-expression.
In both cases, we descend into the expression and check its subexpressions, as they may have nested properties.
Such nested properties are unlikely to be written directly by the user, but might occur after program transformations such as inlining.

Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
First, we evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
% This rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.

We do not generally verify programs directly using the checked semantics.
Instead, the checked semantics is intended as a specification describing what it means to be a valid program.
To verify concrete programs, we generally translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).
% The semantics for contracts is fairly subtle, but we believe it is simpler than the translation to transition systems.
% Having a relatively simple high-level semantics of checks and contracts leaves room for future optimisations in translation to abstract transition systems.
% Additionally, existing contract support in Kind2~\cite{champion2016kind2} has rather ad-hoc restrictions, such as disallowing calling other contracts inside contract relies and guarantees. Our semantics handles this case without any special treatment.

@@

Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
Blessing an expression requires a proof that the checked semantics hold for all input streams:

@@

>>>>

We use an auxiliary function $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$; for primitives $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ denotes that $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

@@

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types.

@@

Rule \textsc{TCheck} states that checked properties $\xcheckP{\PStatus}{e}$ require a boolean property $e$.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee must be boolean expressions, and the guarantee can refer to the result as $x$.

@@

At a high level, evaluation unfolds recursive streams to determine a concrete value.
For example, to evaluate the earlier sum example with input $i = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumionetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream to get $(\xfby{0}{\sumrec}) + i$.
Evaluation of primitives is standard.
For variables, we find the variable in the current heap:
$$
\ruleIN{
}{
  \bigstep{\sumionetwo}{i}{2}
}{Var}
$$

@@

Returning to \autoref{f:core-bigstep},
rule {\sc Var} evalutes a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap.
Rules {\sc Value} and {\sc Let} are standard.
Rule {\sc Prim} evaluates a primitive $p$ applied to many arguments $e_1$ to $e_n$ by evaluating each argument separately; we then apply the primitive with prim-sem metafunction.
% rule {\sc Value} states that evaluating a value results in the value itself.
% Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
% Rule {\sc Let} is standard.

For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value.
When there is no previous value -- the streaming history only contains the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
Otherwise, rule $\textsc{Fby}_S$ applies; we evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.

@@

Rule {\sc Check} ignores the property when evaluating check expressions.
We do not dynamically check the property here; this is done in the checked semantics (\autoref{s:core:checked}).

Similarly, rule {\sc Contract} ignores preconditions and postconditions when evaluating contracts.
From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
However, such an abstraction would make evaluation non-deterministic and, for contracts with unsatisfiable postconditions, non-total.
The deterministic and total nature of evaluation is key to our proofs and metatheory.

@@

In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
We call these the \emph{checked} semantics; they are comparable to checking runtime assertions.
% Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.

The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties and contracts of $e$ with status $\PStatus$ hold.
The property status dictates which properties should be checked and which should be ignored.

We consider a program to be \emph{valid} if its checks hold for all histories ($\forall \Sigma.~ \semcheck{\Sigma}{\PSValid}{e}$).
The checked semantics are a specification describing what it means to be a valid program.
We do not generally verify programs directly using the checked semantics; instead, we translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).

% To show that an expression $e$'s unknown properties hold, we show that for all streaming histories $\Sigma$, the unknown properties hold ($\semcheck{\Sigma}{\PSUnknown}{e}$).
% When performing this proof, we can assume that the valid properties already hold to avoid re-checking properties ($\semcheck{\Sigma}{\PSValid}{e}$).
To check a property $(\xcheckP{\PStatus}{e})$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.

Checking contracts is more involved.
Conceptually, contracts involve two kinds of checks: one for the definition and one for the call-site.
To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
% The fact that the checked semantics refers to a particular $\Sigma$ is significant: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.

@@

For example, recall our earlier contract that the sum of strictly positive integers is positive:
$$
@let@~\text{sum}~i = \sumcontract
$$
To check the contract definition on a concrete input $i = [1; 2]$, we first evaluate the body:
$$
\bigsteps{\sumionetwo}{\sumrec}{[1; 3]}
$$
We then check that, assuming all inputs are positive, then all results are positive:
$$
\bigstepalways{\sumionetwo}{i > 0} \implies
\bigstepalways{\sgl{i \mapsto 1, \sumvar \mapsto 1}; \sgl{i \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
$$

It is crucial that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $i = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.

To check a contract instance, we need to check whether the call-site satisfies the rely.

% Contract instantiations often involve feedback loops, where the input to a contract depends on previous output values of the contract itself.
% For example, we can compute the Fibonacci sequence by instantiating the sum contract with its output fed back to it\footnote{The program shown here does not strictly fit within the core grammar, which does not support named functions; in the implementation we use the \fstar{} meta-language's top-level definitions for such definitions.}:
% $$
%   % @rec@~\fibvar.\, \sumcontractX{\PSUnknown}{(\xfby{1}{\fibvar})}
%   @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
% $$

% To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
% To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
% We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.

% To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
% Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
% This sort of feedback is necessary for proving properties of mutually-dependent calls.
% This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
% For example, the non-causal program $@rec@~\textit{nope}.\, \text{sum}~(-\textit{nope})$ depends instantaneously on itself.
% The rely requires that the negated result is positive; the guarantee ensures that the result is positive.
% However, there is no positive number whose negation is also positive, which suggests that the contract itself is inconsistent.
% This potential issue is resolved by outlawing such non-causal programs.

@@

The checked semantics of Pipit is defined in \autoref{f:core-check}.
% The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
% For binding forms such as let-expressions and recursive-expressions, the semantics evaluates the corresponding expression and extends the streaming history with the bound variable.

@@

Rule {\sc ChkCheck} checks the properties marked $\PStatus$ in an expression $\xcheckP{\PStatus'}{e}$.
If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we evaluate the expression $e$ and require it to be true at all steps.
We then unconditionally descend into the subexpression to check any nested properties.
Such nested properties are unlikely to be written directly by the user, but might occur after inlining.

Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
This rule checks both the contract definition and the call-site.
We evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
Although the contract only has one property status, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition (assumed to be $\PSValid$).
To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
% This rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.

% The semantics for contracts is fairly subtle, but we believe it is simpler than the translation to transition systems.
% Having a relatively simple high-level semantics of checks and contracts leaves room for future optimisations in translation to abstract transition systems.
% Additionally, existing contract support in Kind2~\cite{champion2016kind2} has rather ad-hoc restrictions, such as disallowing calling other contracts inside contract relies and guarantees. Our semantics handles this case without any special treatment.

@@

Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
% Blessing an expression requires a proof that the unknown checks hold:
Blessing an expression requires a proof that, for all input streams, assuming the valid checks hold, then the unknown checks hold:

@@

We generally prove these properties via abstract transition systems (\autoref{s:transition}).

### 2024-03-28 2024pipit/section/40-Transition.tex
copy and tighten

For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: ``there exists some value that satisfies the specification''.
Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.

@@

To verify the contract definition, we first translate it to an abstract transition system whose input environment contains an integer \emph{ints}, and whose result type is also an integer.
The followed-by delay results in a local state variable called sum_fby, and the recursive stream results in a free variable called sum.
We get the following transition system:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> free  \> = (sum: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ sum_fby = f.sum \}; \\
  \> \> \> value  \> = f.sum; \\
  \> \> \> rely   \> = f.sum = (s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
  \> \> \> guar   \> = f.sum > 0; \} \\
  \} \\
  \end{tabbing}

The initial state of 0 corresponds to the initial value of the followed-by.
In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containins f.sum, and $s$ refers to the state heap containing s.sum_fby.
The result of the recursive stream is referred to as f.sum; in the rely / assumption of the step result, f.sum is restricted to have the correct sum.
The step result's rely also includes the contract's rely, which allows proofs of system validity to assume that the input integer is positive.
Finally, the step result's guarantee includes the contract's guarantee, which ensures that the proof of system validity actually proves that the contract holds.

The translation for contract instantiations proceeds similarly, except that the contract itself is replaced by an arbitrary value in the free context.
For example, recall the Fibonacci sequence:
$
  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
$.

This program does not require any input values, so we leave the input context empty.
The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
Similarly, the free context includes an entry for the recursive stream, and an entry for the contract itself:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ fib_def: system () $\ZZ$ = \{ \\
  \> state   \> = (fib_fby: $\ZZ$); \\
  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
  \> init  \> = \{ fib_fby = 1 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ fib_fby = f.fib \}; \\
  \> \> \> value  \> = f.fib; \\
  \> \> \> rely   \> = f.fib = f.sum_contract $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
  \> \> \> guar   \> = s.fib_fby > 0; \} \\
  \} \\
  \end{tabbing}

@@

Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
This minor optimisation is sound, as the proof of system validity requires all steps to satisfy the rely and guarantee; we discuss this further in \autoref{s:transition:ind}.

@@

Our definition of induction over a transition system's inputs is somewhat non-standard, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i: \textit{input}) (f: \textit{sys}\text{.free}).$ \\
  \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
\end{tabbing}

@@

This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
K-induction is a fairly standard invariant strengthening technique~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.

@@

As most expressions do not modify the state heap, the invariant for most expressions simply descends into the subexpressions.
Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.

>>>>

For recursive streams and contract instantiations, which hide their implementation, the natural translation to a transition system would involve existentially quantifying a result that satisfies the specification.
Unfortunately, using an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.

@@

To verify the contract definition, we first translate it to an abstract transition system whose input environment contains an integer \emph{ints}, and whose result type is also an integer.
The followed-by delay results in a local state variable called sum_fby, and we encode the existentially-quantified recursive stream as a free context variable called sum:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> free  \> = (sum: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ sum_fby = f.sum \}; \\
  \> \> \> value  \> = f.sum; \\
  \> \> \> rely   \> = (f.sum = s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
  \> \> \> guar   \> = f.sum > 0; \} \}
  \end{tabbing}

The initial state of 0 corresponds to the initial value of the followed-by.
In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containing the recursive stream f.sum, and $s$ refers to the state heap containing s.sum_fby.
In the rely of the step result, f.sum is constrained to match the body of the recursive stream.
The step result's rely also includes the contract's rely that the input integer is positive.
Finally, the step result's guarantee includes the contract's guarantee that the contract holds.

To verify the transition system, we prove inductively that if the rely always holds, then the guarantee holds; we discuss proofs of system validity further in \autoref{s:transition:ind}.

The translation for contract instantiations is similar, except that the contract body is replaced by an arbitrary value from the free context.
For example, we can use the sum contract to implement the Fibonacci sequence with
$
  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
$.
This program does not require any input values, so we leave the input context empty.
The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
Similarly, the free context includes an entry for the recursive stream, and an entry for the contract itself:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ fib_def: system () $\ZZ$ = \{ \\
  \> state   \> = (fib_fby: $\ZZ$); \\
  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
  \> init  \> = \{ fib_fby = 1 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ fib_fby = f.fib \}; \\
  \> \> \> value  \> = f.fib; \\
  \> \> \> rely   \> = (f.fib = f.sum_contract) \\
  \> \> \>        \> $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
  \> \> \> guar   \> = s.fib_fby > 0; \} \}
  \end{tabbing}

@@

Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
This minor optimisation is sound, as we define system validity to require all steps to satisfy the rely.

@@

Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i: \textit{input}) (f: \textit{sys}\text{.free}).$ \\
  \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
\end{tabbing}

@@

This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.

@@

As most expressions do not modify the state heap, the invariant for most expressions simply descends into the subexpressions.
Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
With this invariant, we can prove abstraction:

### 2024-03-28 2024pipit/section/50-Extraction.tex
copy and tighten

Here, the step function takes records describing instances of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value.
This duplication of work is often removed by partial evaluation and dead-code-elimination, which are performed during code extraction.

This translation to transition systems is verified to preserve the original semantics.
The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we rely on the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).

@@

To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
This specialisation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.

@@

Unfortunately, our current approach is unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
% We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

>>>>

Here, the step function takes records describing instances of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value.
This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.

This translation to transition systems is verified to preserve the original semantics.
The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we use the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).

@@

To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to partially-evaluate the application of the translation to a particular input program.
This partial-evaluation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.

@@

Unfortunately, our current approach is unsuitable for generating imperative array code, as our pure transition system only supports pure arrays.
In the future, we intend to support efficient array computations and fix the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
% We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.
